---
title: "Ungulate cestode life history - Propagules"
author: "Dan Benesh"
output: html_document
---

I use the life cycle database to explore ideas about the costs and benefits of different infection routes exhibited by helminths of grazing mammals (ungulates). In this document, I specifically focus on cestode propagules. In essence, I compare two taxa, (i) taeniids that are egg-transmitted and use ungulates as intermediate hosts and (ii) anoplocephalids that are transmitted to ungulates by mite intermediate hosts. My main goal is to see if patterns in nematodes are also observed in cestodes.

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ape)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE)
theme_set(new = theme_bw())
```

```{r importdata}
hosts <- read.csv(file = "../../data/CLC_database_hosts.csv", header = TRUE) 
lh <- read.csv(file = "../../data/CLC_database_lifehistory.csv", header = TRUE)
host_mass <- read.csv(file = "../../data/collated_host_mass_data.csv", header = TRUE)
host_tax <- read.csv(file = "../../data/ncbi_host_taxonomy.csv", header = TRUE)
```
```{r}
# filter out ungulate host species
ungulate_host_species <- filter(host_tax, order == "Perissodactyla" | order == "Laurasiatheria")%>%
  select(sp.query, phylum, class, order, family, genus)
```
```{r}
# identify ungulates in list of hosts, add var
hosts <- mutate(hosts, 
                ungulate = if_else(Host.species %in% ungulate_host_species$sp.query, 1, 0))
# print common names
# filter(hosts, ungulate == 1)%>%select(Host.species, Host.common.name)%>%distinct()%>%
#   knitr::kable(format = 'html')
```
```{r}
# sp by lc data table
lcl <- group_by(hosts, Parasite.species, Parasite.group)%>%
  summarise(lcl = max(Host.no))
```
```{r}
# calc biomasses
lh <- mutate(lh, biovolume = 
                  if_else(Shape %in% c("cylinder", "thread-like", "whip"), 
                          pi * (Width/2)^2 * Length, # calculate volume as a cylinder
                          if_else(Shape %in% c("coiled", "sphere", "ellipsoid"),
                                  4/3 * pi * Length/2 * Width/4, # calculate volume as a ellipsoid
                                  Length * Width # calculate volume as area for remaining ribbon, leaf shapes
                                  )),
                biovolume = biovolume * 1.1) # covert to biomass with assumed 1.1. g/cm3 tissue density 

# species averages at adult stage
lh_biov_sp <-filter(lh, Stage == 'adult', (Sex == 'f' | is.na(Sex)) )%>% # remove adult males
  group_by(Parasite.species)%>%
  summarize(adult_biov = mean(biovolume, na.rm=T),
            adult_dt = mean(Development.time, na.rm = T))
```

```{r}
# calc mean host sizes
host_mass_avg <- group_by(host_mass, binomial)%>%
  summarize(body.mass = mean(body.mass, na.rm=T))
# add to hosts db
hosts <- left_join(hosts, host_mass_avg, by = c("Host.species" = "binomial"))
```
```{r}
# filter to ungulate-infecting helminths, excluding those accidentally or atypically found in ungulates
ung_parasites <- filter(hosts, ungulate == 1, Typical.host == 'typical', Parasite.group == "cestode")%>%
  group_by(Parasite.species, Host.no, Def.int)%>%
  summarize(host.bm = mean(body.mass, na.rm = T))
```
```{r}
# combine parasite sizes, host sizes, life cycle info for plotting
ung_parasites <- left_join(ung_parasites, lcl)
ung_parasites <- left_join(ung_parasites, lh_biov_sp)
ung_parasites <- mutate(ung_parasites, complex = if_else(lcl == 1, "direct", "complex"))

# add hatching strategy to plot data
hatch <- select(lh, Parasite.species, hatch = Egg.hatch)%>%
  mutate(hatch = if_else(hatch == 'eaten', 'egg eaten', 'hatched larva\neaten'))%>% # simplify hatch strategy
           na.omit()%>%distinct()
ung_parasites <- left_join(ung_parasites, hatch)

# make var combining hatching and life cycle strategies
ung_parasites <- mutate(ung_parasites, 
                        trans_strat = if_else(Host.no == 2, "intermediate host\neaten", hatch))
```
```{r}
# filter propagule sizes, average them within species
prop_sizes <- filter(lh, Host.no == 0)%>%
  select(Parasite.species, Stage, biovolume)%>%
  group_by(Parasite.species, Stage)%>%
  summarize(propagule_size = mean(biovolume, na.rm = T))
```
```{r}
# make variable for selecting transmission propagules eaten
hatch$selector <- "free larva"
hatch$selector[which(hatch$hatch == "egg eaten")] <- "egg"

# make variable for selecting size at initiation of growth after infection
hatch$selector2 <- "free larva"
hatch$selector2[which(hatch$hatch == "egg eaten")] <- "embryo"
```
```{r}
# reduce to just the transmission propagules to the ungulates
prop_sizes <- left_join(prop_sizes, hatch, by = "Parasite.species")
prop_sizes_eaten <- filter(prop_sizes, selector == Stage)%>%
  select(Parasite.species, propagule_eaten = Stage, propagule_size_eaten = propagule_size)
prop_sizes_growing <- filter(prop_sizes, selector2 == Stage)%>%
  select(Parasite.species, propagule_growth_start = Stage, propagule_size_grow_start = propagule_size)
prop_sizes <- left_join(prop_sizes_eaten, prop_sizes_growing)
prop_sizes <- filter(prop_sizes, Parasite.species %in% ung_parasites$Parasite.species)
```
```{r}
ung_parasites <- left_join(ung_parasites, prop_sizes)
```

We can start by comparing the propagule sizes of these two groups. The bigger eggs belong to anoplocephalids, the worms targeting mites as next host.

```{r}
ggplot(ung_parasites, aes(x = trans_strat, y = propagule_size_eaten)) + 
  geom_boxplot() +
  geom_point(position = position_jitter(width = .2, height = 0)) +
  scale_y_log10()
```

This is not the pattern seen in nematodes. In nematodes, egg-to-ungulate species had larger propagules than species transmitted to an intermediate host. A couple differences may be relevant. First, egg-to-ungulate cestodes (Taeniids) are transmitted via carnivore faeces, whereas the egg-transmitted nematodes have direct cycles and are thus transmitted by ungulate faeces. Faecal avoidance behaviors and the trophic vacuum they cause may differ between these groups. Second, cestode eggs are often packed within full proglottids, which may or may not disintegrate upon release into the environment. In other words, it is unclear if the size of a single egg or the size of a proglottid best represents the propagule consumed by the target host.

When we look at embryo size within eggs (oncospheres), we see that they are quite comparable. 

```{r}
ggplot(ung_parasites, aes(x = trans_strat, y = propagule_size_grow_start)) + 
  geom_boxplot() +
  geom_point(position = position_jitter(width = .2, height = 0)) +
  scale_y_log10()
```

Both groups belong to Cyclophyllidea and I believe cyclophyllideans have rather similar oncospheres. Apparently, the eggs of anoplocephalids have a thicker shell than those of taeniids, causing their larger overall size. A quick google image search confirms this, e.g. the large, irregular shell is even diagnostic for *Moniezia*, the best studied anoplocephalid.

There is a bit of a tendency for larger eggs to harbor larger embryos in both groups, though the ratio is different, because anoplocephalids (blue) have large shells relative to the size of the embryo.

```{r}
ggplot(ung_parasites, aes(x = propagule_size_eaten, y = propagule_size_grow_start, color = trans_strat)) + 
  geom_point(position = position_jitter(width = .2, height = 0)) +
  scale_y_log10() + scale_x_log10() + geom_smooth(se = F, method = 'lm')
```

In nematodes, we found an association between transmission strategy and propagule development - propagules targeting ungulates developed in the external environment, while those targeting intermediate hosts did not. We could not confirm this pattern in cestodes. Both taeniids and anoplocephalids are released into the external environment infective (i.e. the eggs are embryonated).

```{r}
temp_test <- filter(lh, Parasite.group == 'cestode', Host.no == 0)%>%
  select(Parasite.species, Stage, Egg.hatch, Development.time, Temp, Author)%>%
  mutate(degree.days = Development.time*(Temp - 5))%>%
  filter(!is.na(Development.time))
```

Some cestodes require egg development and hatching before transmission to the first host. When we isolate the cestode species with data for propagule developmental time, we see that they all target aquatic first hosts.

```{r}
knitr::kable(
  filter(hosts, Parasite.species %in% temp_test$Parasite.species, Host.no == 1)%>%
  select(Parasite.species, Host.no, Host.common.name, Host.habitat)%>%distinct()
)
```

Thus, I am unaware of any tapeworm species that releases unembryonated eggs into a terrestrial environment that require development before becoming infective.

# Conclusions

Patterns of propagule size and development are not the same in ungulate-infecting cestodes and nematodes. Cestodes targeting invertebrate intermediate hosts have larger propagules than those targeting ungulates, which is not the same pattern observed in nematodes. Nematode transmission strategies are related to their propagule development, whereas in cestodes there is simply no variation in propagule development (i.e. they are all embryonated). It is unclear if this incongruence is caused by ecological differences in transmission (e.g. eggs in carnivore faeces in taeniids instead of ungulate faeces) or by phylogenetic constraints (e.g. propagule development being less constrained in nematodes).