---
title: "Getting taxonomic tree to map invasion of ungulates"
author: "Dan Benesh"
output: html_document
---

In this script, I get a tree from the OTL that will be used to map ungulate parasitism across helminths. I want to know how often these hosts have been invaded and which are most commonly strategies used to do so.

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(taxize)
library(rotl)
library(ape)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE)
theme_set(new = theme_bw())
```

First, I make a list of helminth families in which ungulate parasitism occurs. I also listed the transmission routes for each of these families. I defined 6 routes to ungulates - the first three are trophic (ungulate eats propagule), the last three are not:
1. Eggs are ingested by ungulate
2. Larvae on herbage eaten (free L3s in nematodes, encysted metacercariae in trematodes)
3. Intermediate host ingested by ungulate
4. Larvae actively seek host, usually penetrate it
5. Vector-transmission
6. No free propagule stage - transmammary, transplacental, or cannibalistic transmission

The last group is kind of a miscellaneous category, where none of the strategies are widespread. Some worms do not fit neatly into these categories. For instance, *Toxocaris* is a nematode that, like most ascarids, is transmitted through ingestion of eggs. However, it can also be transmitted from cow to calf via the milk. Thus, it could be assigned two 'states' simultaneously (1 or 6). Similarly, *Strongyloides* can be transmitted vertically from mother to offspring, but it also produces free L3 larvae that penetrate skin (4 or 6). And in one final case, the pig kidney worm *Stephanurus* (Syngamidae) can be transmitted via skin penetration (unusual in the species), ingestion of free larvae, or via a paratenic earthworm host (probably the most common case) (routes 2, 3, or 4 exist here). Since I am interested in mapping how many times different transmission strategies evolved, it may be worthwhile to look at the phylogeny, see what the relatives of these flexible species are doing, and then assign them a strategy to ensure the novel innovations in these taxa are counted.

Families and transmission styles were compiled by Jimmy Chubb.

```{r}
# list of helminth fams in ungulates - from MS
mon_ung_fams <- "Polystomatidae"
ces_ung_fams <- c('Diphyllobothridae', 'Anoplocephalidae', 'Linstowiidae', 'Thysanosomatinae', 'Inermicapsiferinae',
              'Davaineidae', 'Taeniidae')
tre_ung_fams <- c('Hasstilesiidae', 'Schistosomatidae', 'Echinostomatidae', 'Fasciolidae', 'Paramphistomidae',
              'Balanorchiidae', 'Brumptiidae', 'Choerocotyloididae', 'Cladorchiidae', 'Gastrodiscidae',
              'Gastrothylacidae', 'Olveriidae', 'Stephanopharyngidae', 'Zygocotylidae', 'Notocotylidae',
              'Opisthorchiidae', 'Heterophyidae', 'Dicrocoeliidae', 'Paragonimidae')
nem_ung_fams <- c('Capillariidae', 'Robertdollfusidae', 'Trichinellidae', 'Trichuridae', 'Ancylostomatidae',
               'Ascarididae', 'Toxocaridae', 'Chabertiidae', 'Cooperiidae', 'Dictyocaulidae', 'Filariidae',
               'Gnathostomatidae', 'Gongylonematidae', 'Habronematidae', 'Haemonchidae', 'Metastrongylidae',
               'Molineidae', 'Onchocercidae', 'Oxyuridae', 'Protostrongylidae', 'Spirocercidae', 'Strongylidae',
               'Strongyloididae', 'Syngamidae', 'Thelaziidae', 'Trichostrongylidae')
aca_ung_fams <- c('Oligacanthorhynchidae','Echinorhynchidae')
```
```{r}
route1 <- "eggs ingested"
route2 <- "larvae on herbage"
route3 <- "intermediate host eaten"
route4 <- "larvae seek/penetrate"
route5 <- "vector-transmission"
route6 <- "no free propagule"
trans_routes <- data.frame(rbind(c("Polystomatidae", route4),
                                 c("Diphyllobothridae", route3),
                                 c('Anoplocephalidae', route3),
                                 c('Linstowiidae', route3),
                                 c('Davaineidae', route3),
                                 c('Taeniidae', route1),
                                 c('Hasstilesiidae', route3),
                                 c('Schistosomatidae', route4),
                                 c('Echinostomatidae', route3),
                                 c('Fasciolidae', route2),
                                 c('Paramphistomidae',route2),
                                 c('Balanorchiidae', route2),
                                 c('Brumptiidae', route2),
                                 c('Choerocotyloididae', route2),
                                 c('Cladorchiidae', route2),
                                 c('Gastrodiscidae', route2),
                                 c('Gastrothylacidae', route2),
                                 c('Olveriidae', route2),
                                 c('Stephanopharyngidae', route2),
                                 c('Zygocotylidae', route2),
                                 c('Notocotylidae', route2),
                                 c('Opisthorchiidae', route3),
                                 c('Heterophyidae', route3),
                                 c('Dicrocoeliidae', route3),
                                 c('Paragonimidae', route3),
                                 c('Capillariidae', route1),
                                 c('Robertdollfusidae', route5),
                                 c('Trichinellidae', route6),
                                 c('Trichuridae', route1),
                                 c('Ancylostomatidae', route4),
                                 c('Ascarididae', route1),
                                 c('Toxocaridae', route1),
                                 c('Chabertiidae', route2),
                                 c('Cooperiidae', route2),
                                 c('Dictyocaulidae', route2),
                                 c('Filariidae', route5),
                                 c('Gnathostomatidae', route3),
                                 c('Gongylonematidae', route3),
                                 c('Habronematidae', route5),
                                 c('Haemonchidae', route2),
                                 c('Metastrongylidae', route3),
                                 c('Molineidae', route2),
                                 c('Onchocercidae', route5),
                                 c('Oxyuridae', route1),
                                 c('Protostrongylidae', route3),
                                 c('Spirocercidae', route3),
                                 c('Strongylidae', route2),
                                 c('Strongyloididae', route4),
                                 c('Syngamidae', route3),
                                 c('Thelaziidae', route5),
                                 c('Trichostrongylidae', route2),
                                 c('Oligacanthorhynchidae', route3),
                                 c('Echinorhynchidae', route3)
))
names(trans_routes) <- c('family', 'trans_route')
rm(route1, route2, route3, route4, route5, route6)
```

A table of the different transmission routes shows that 1) trophic transmission dominates (`r 6+18+19` families versus `r 4+1+5` families with non-trophic infection routes), and 2) egg ingestion is the least common trophic route, consistent with the idea of a vacuum caused by faecal avoidance.

```{r}
knitr::kable(table(trans_routes$trans_route))
```

Instead of counting families, though, we want to count the independent evolutionary transitions to helminth parasitism. To be able to map these ancestral states, we also need to know in which helminths ungulate parasitism does not occur. In other words, we need to get the related taxa that are not parasites of ungulates. For this, I queried COL and ITIS with `taxize` to find all families in Acanthocephala, Nematoda, and Platyhelmintha.

Starting with the acanths, who have only sporadically been found in ungulates. Both families with records from ungulates were in the taxonomy databases.

```{r}
# from COL
aca_col_id <- "29cf5a3a0cad337a0216edca32bab745" # fetched before
aca_col_fams <- downstream(aca_col_id, downto = 'family', db = 'col')
aca_col_fams <- aca_col_fams$`29cf5a3a0cad337a0216edca32bab745`$childtaxa_name
# from ITIS
aca_itis_id <- 64238 # looked it up online
aca_itis_fams <- downstream(aca_itis_id, downto = 'family', db = 'itis')
aca_itis_fams <- aca_itis_fams$`64238`$taxonname

#sum(aca_col_fams %in% aca_itis_fams)/length(aca_col_fams) # percent COL in ITIS
#sum(aca_ung_fams %in% aca_col_fams)/length(aca_ung_fams) # percent helminth fams in COL
#sum(aca_ung_fams %in% aca_itis_fams)/length(aca_ung_fams) # percent helminth fams in ITIS
```

Twenty-five nematode families have been recorded in ungulate hosts. About 80% of these were in COL, but only about 50% in ITIS.

```{r}
# get list of all nematode families
# from COL
nem_col_id <- get_colid(sciname = "Nematoda")
nem_col_fams <- downstream(nem_col_id, downto = 'family', db = 'col')
nem_col_fams <- nem_col_fams$`562405d59c95f93423494a00939ae2ff`$childtaxa_name
# from ITIS
nem_itis_id <- get_tsn(searchterm = "Nematoda")
nem_itis_fams <- downstream(nem_itis_id, downto = 'family', db = 'itis')
nem_itis_fams <- nem_itis_fams$`59490`$taxonname

#sum(nem_col_fams %in% nem_itis_fams)/length(nem_col_fams) # percent COL in ITIS
#sum(nem_ung_fams %in% nem_col_fams)/length(nem_ung_fams) # percent helminth fams in COL
#sum(nem_ung_fams %in% nem_itis_fams)/length(nem_ung_fams) # percent helminth fams in ITIS
```

Each of the three groups of parasitic playhelminths (Monogenea, Trematoda, Cestoda) have members infecting ungulates. But only about 40 to 60% of these 27 families have a match in the two taxonomy databases. The discrepancy suggests either Jimmy or the databases are using outdated family names, which is a common issue.

```{r}
# get list of all platy families
# from COL
pla_col_id <- get_colid(sciname = "Platyhelminthes")
pla_col_fams <- downstream(pla_col_id, downto = 'family', db = 'col')
pla_col_fams <- pla_col_fams$`7253b0d111fd5eb0c6a9856e0d8d1cad`$childtaxa_name
# from ITIS
pla_itis_id <- get_tsn(searchterm = "Platyhelminthes")
pla_itis_fams <- downstream(pla_itis_id, downto = 'family', db = 'itis')
pla_itis_fams <- pla_itis_fams$`53963`$taxonname

pla_ung_fams <- c(mon_ung_fams, tre_ung_fams, ces_ung_fams)
#sum(pla_col_fams %in% pla_itis_fams)/length(pla_col_fams) # percent COL in ITIS
#sum(pla_ung_fams %in% pla_col_fams)/length(pla_ung_fams) # percent helminth fams in COL
#sum(pla_ung_fams %in% pla_itis_fams)/length(pla_ung_fams) # percent helminth fams in ITIS
```

Given that the helminth ungulate families and the taxonomy datasets are not completely congruent, I chose to Combine all three lists of family names to query the OTL. Some might be redundant or invalid, but this should give the largest, most complete tree with which to map ancestral states.

```{r}
# combine theme all
aca_fams_all <- unique(c(aca_ung_fams, aca_col_fams, aca_itis_fams))
nem_fams_all <- unique(c(nem_ung_fams, nem_col_fams, nem_itis_fams))
pla_fams_all <- unique(c(pla_ung_fams, pla_col_fams, pla_itis_fams))
```

```{r}
# search OTL db
aca_search <- tnrs_match_names(names = aca_fams_all, context_name = "Animals")
nem_search <- tnrs_match_names(names = nem_fams_all, context_name = "Nematodes")
pla_search1 <- tnrs_match_names(names = pla_fams_all[1:240], context_name = "Platyhelminthes")
pla_search2 <- tnrs_match_names(names = pla_fams_all[241:length(pla_fams_all)], context_name = "Platyhelminthes")
```

```{r}
# which ones are in tree
aca_ott_in_tree <- ott_id(aca_search)[is_in_tree(ott_id(aca_search))]
nem_ott_in_tree <- ott_id(nem_search)[is_in_tree(ott_id(nem_search))]
pla_ott <- c(ott_id(pla_search1), ott_id(pla_search2))
pla_ott_in_tree <- pla_ott[is_in_tree(pla_ott)]
# attributes erased by appending lists together - re-add so subtree function works
attr(pla_ott_in_tree, 'class') <- c("otl_ott_id", "list")
```

```{r}
# get subtrees
aca_tree <- tol_induced_subtree(ott_ids = aca_ott_in_tree, label_format = 'name')
nem_tree <- tol_induced_subtree(ott_ids = nem_ott_in_tree, label_format = 'name')
pla_tree <- tol_induced_subtree(ott_ids = pla_ott_in_tree, label_format = 'name')
```

Of `r length(aca_fams_all) + length(nem_fams_all) + length(pla_fams_all)` families queried, `r length(aca_tree$tip.label) + length(nem_tree$tip.label) + length(pla_tree$tip.label)` were included in the tree. All acanth families of ungulates were in the tree. Here are the nematodes families of ungulates that were not included.

```{r}
# ungulate nem families not in nem tree
nem_ung_fams[which(!(nem_ung_fams %in% nem_tree$tip.label))]
```

And here are the platyhelminth families infecting ungulates not in the tree.

```{r}
pla_ung_fams[which(!(pla_ung_fams %in% pla_tree$tip.label))]
```

Some of these do not need to be added to the tree, such as the subfamilies of Anoplocephalidae (Linstowiinae, Thysanosomatinae, Inermicapsiferinae), because a similar type of transmission route applies to all these families (mite transmission). In other cases, the families are in the tree, but their name was changed, e.g. Diphyllobothridae to Diphyllobothriidae. For other cases, a closer look at the tree is warranted.

I bound the trees together to create a single tree for the helminth taxa.

```{r}
aca_tree$root.edge <- .2 # root edge needed to bind trees not exactly at root
nem_tree$root.edge <- .2
pla_tree$root.edge <- .2
comb_tree <- bind.tree(pla_tree, aca_tree, position = .2)
comb_tree <- bind.tree(comb_tree, nem_tree, position = .2)
```

```{r}
# make dataset
helminth_ungs <- data.frame(family = c(aca_fams_all, nem_fams_all, pla_fams_all))
helminth_ungs <- mutate(helminth_ungs, group = if_else(family %in% aca_fams_all, "Acanthocephala",
                                                       if_else(family %in% pla_fams_all, "Platyhelminthes",
                                                               if_else(family %in% nem_fams_all, "Nematoda", '?'))),
                        ung_inf = family %in% c(aca_ung_fams, nem_ung_fams, pla_ung_fams))

# add names from OTL into data
search_all <- rbind(aca_search, nem_search, pla_search1, pla_search2)
mv <- match(tolower(helminth_ungs$family), search_all$search_string)
helminth_ungs$otl_name <- search_all$unique_name[mv]
helminth_ungs <- mutate(helminth_ungs, otl_name = gsub(" ", "_", otl_name)) # replace spaces with underscores to match tip labels

# add transmission route
helminth_ungs <- left_join(helminth_ungs, trans_routes, by = 'family')
helminth_ungs <- mutate(helminth_ungs,
                        trans_route = if_else(is.na(trans_route), "not ungulate parasites", trans_route))

# which ones are in tree?
helminth_ungs <- mutate(helminth_ungs, in_tree = otl_name %in% comb_tree$tip.label)

# remove 'not assigned' row
helminth_ungs <- filter(helminth_ungs, family != 'Not assigned')
```

Here are the helminth taxa of ungulates that were not included in the tree.

```{r}
filter(helminth_ungs, ung_inf, !in_tree)
```

A couple can be ignored - Thysanosomatinae and Inermicapsiferinae are subfamilies of Anoplocephalidae, which is in the tree and exhibits no variation in transmission strategy. The nematode groups, like Ancylostomatidae and Ascarididae, are well studied and their exclusion probably represents evolving taxonomy. They can very likely be added manually. The 3 trematodes (Balanorchiidae, Choerocotyloididae, and Zygocotylidae) are closely related (same superfamily). Given they exhibit the same transmission strategy, their inclusion or exclusion may have no effect on the evolutionary counts.

I wrote two files - a nexus file with the tree and a csv with data on the families queried here.

```{r}
# write files
write.csv(helminth_ungs, file = "ungulate_helminth_tree_table.csv", row.names = F)
write.nexus(comb_tree, file = "ungulate_helminth_tree.nex")
```


```{r}
# list of helminth taxa in ungulates - from London NHM DB, need ungulates for cross-checking
```
