---
title: "Improve tree"
author: "Dan Benesh"
output: html_document
---

This document improves the tree output by the [get tree](getting_taxonomy_helminth_tree.Rmd) notebook.

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ape)
library(castor) # for ancestral state mapping
library(ggtree)
library(RColorBrewer)
library(phytools)
library(geiger)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE)
theme_set(new = theme_bw())
```

```{r get tree}
tree <- read.nexus(file = "ungulate_helminth_tree.nex")
tree$edge.length <- NULL # remove branch lengths
```

```{r get data}
tre_dat <- read.csv(file = "ungulate_helminth_tree_table.csv")
```

I wrote a function that takes in a tree and the associated data and outputs a figure with ancestral states mapped onto it. I used the function to iteratively improve the tree and to visualize the improvements.

```{r}
source("tree_checking_function.R")
```

Make the names in the data and tree consistent

```{r}
# in tree
fix_tip_names <- substr(tree$tip.label, start = 1, stop = regexpr(tree$tip.label, pattern = "_") - 1 )
tree$tip.label[which(fix_tip_names!="")] <- fix_tip_names[which(fix_tip_names!="")]

fix_node_names <- substr(tree$node.label, start = 1, stop = regexpr(tree$node.label, pattern = "_") - 1 )
tree$node.label[which(fix_node_names!="")] <- fix_node_names[which(fix_node_names!="")]

# in data
fix_data_names <- substr(tre_dat$otl_name, start = 1, stop = regexpr(tre_dat$otl_name, pattern = "_") - 1 )
tre_dat$otl_name_red <- tre_dat$otl_name
tre_dat$otl_name_red[which(fix_data_names!="")] <- fix_data_names[which(fix_data_names!="")]

rm(fix_data_names, fix_node_names, fix_tip_names)
```


```{r}
tree_plot <- tree_checking(tree, tre_dat)

#ggsave(filename = 'for_resolving.pdf', width = 500, height = 400, units = "mm")
```

The first needed step is to **add missing tips** to the tree. The following nematode families were added: Ancylostomatidae, Ascarididae, Trichostrongylidae.

```{r}
tree <- compute.brlen(tree, power = 1 ) # add branch lengths, so taxa can be add along branch, not just at node

# add taxa to branch below internal node
tree <- bind.tip(tree, tip.label = "Ancylostomatidae", 
                 position = 0.0001,
                 where = length(tree$tip.label) + which(tree$node.label == "mrcaott46769ott67070"))
tre_dat$in_tree[which(tre_dat$family == 'Ancylostomatidae')] <- TRUE

# add to tip
tree <- bind.tip(tree, tip.label = "Ascarididae", 
                 where = which(tree$tip.label == "Toxocaridae"))
tre_dat$in_tree[which(tre_dat$family == 'Ascarididae')] <- TRUE


tree <- bind.tip(tree, tip.label = "Trichostrongylidae", 
                 where = which(tree$tip.label == "Haemonchidae"))
tre_dat$in_tree[which(tre_dat$family == 'Trichostrongylidae')] <- TRUE
```

Then the following trematode tips are grafted to the tree: Balanorchiidae, Choerocotyloididae, Zygocotylidae.

```{r}
# add taxa to internal node
tree <- bind.tip(tree, tip.label = "Balanorchiidae", 
                 where = length(tree$tip.label) + which(tree$node.label == "Paramphistomoidea"))
tre_dat$in_tree[which(tre_dat$family == 'Balanorchiidae')] <- TRUE

tree <- bind.tip(tree, tip.label = "Choerocotylidae", 
                 where = length(tree$tip.label) + which(tree$node.label == "Paramphistomoidea"))
tre_dat$in_tree[which(tre_dat$otl_name == 'Choerocotylidae')] <- TRUE

tree <- bind.tip(tree, tip.label = "Zygocotylidae", 
                 where = length(tree$tip.label) + which(tree$node.label == "Paramphistomoidea"))
tre_dat$in_tree[which(tre_dat$family == 'Zygocotylidae')] <- TRUE
```

After adding these families, all the taxa listed in the manuscript's Table 1 are included in the helminth tree. The next step is to add an additional tip for the families exhibiting **multiple strategies** to infect ungulates. For example, the ascarids can be transmitted via egg or transplacentally. Two tips are needed in this case so as to record each strategy as an evolutionary 'event'. 

Here are the groups to which this applies:
Syngamidae (via intermediate host or hatched larvae)
Ascarididae & Toxocaridae (via eggs or trans-placentally, -maternally)
Strongyloididae (via larval penetration or trans-placentally, -maternally)

```{r}
# add to tip
tree <- bind.tip(tree, tip.label = "Syngamidae2", 
                 where = which(tree$tip.label == "Syngamidae"))

tree <- bind.tip(tree, tip.label = "Strongyloididae2", 
                 where = which(tree$tip.label == "Strongyloididae"))

tree <- bind.tip(tree, tip.label = "Ascarididae2", 
                 position = 0.0001,
                 where = getMRCA(tree, tip = c("Ascarididae", "Toxocaridae")))

tree <- bind.tip(tree, tip.label = "Toxocaridae2", 
                 where = which(tree$tip.label == "Ascarididae2"))
```

```{r}
# update data
syg2 <- data.frame(family = 'Syngamidae2', group = 'Nematoda', ung_inf = TRUE,
                   otl_name = 'Syngamidae2', trans_route = 'larvae on herbage', 
                   in_tree = TRUE, otl_name_red = 'Syngamidae2')

str2 <- data.frame(family = 'Strongyloididae2', group = 'Nematoda', ung_inf = TRUE,
                   otl_name = 'Strongyloididae2', trans_route = 'no free propagule', 
                   in_tree = TRUE, otl_name_red = 'Strongyloididae2')

asc2 <- data.frame(family = 'Ascarididae2', group = 'Nematoda', ung_inf = TRUE,
                   otl_name = 'Ascarididae2', trans_route = 'no free propagule', 
                   in_tree = TRUE, otl_name_red = 'Ascarididae2')

tox2 <- data.frame(family = 'Toxocaridae2', group = 'Nematoda', ung_inf = TRUE,
                   otl_name = 'Toxocaridae2', trans_route = 'no free propagule', 
                   in_tree = TRUE, otl_name_red = 'Toxocaridae2')

tre_dat <- bind_rows(tre_dat, syg2, str2, asc2, tox2)
rm(asc2, str2, syg2, tox2)
```

Next, I decided to **remove the clades** of free-living flatworms. These have no effect on the mapping, as they are not parasites, and they make the figures more cluttered.

```{r}
rem_platy <- tips(tree, length(tree$tip.label) + which(tree$node.label == 'Rhabditophora'))
rem_platy2 <- tips(tree, length(tree$tip.label) + which(tree$node.label == 'Catenulida'))
```

```{r}
# drop tips
tree <- drop.tip(tree, tip = rem_platy)
tree <- drop.tip(tree, tip = rem_platy2)
# revise data
tre_dat$in_tree[which(tre_dat$otl_name_red %in%  c(rem_platy, rem_platy2))] <- FALSE
```

```{r}
#tree$edge.length <- NULL # remove branch lengths
#tree_plot <- tree_checking(tree, tre_dat)
#tree_plot
#ggsave(filename = 'for_resolving.pdf', width = 500, height = 400, units = "mm")
```

The improved tree and associated data file were exported.

```{r}
#write.nexus(tree, file = "ungulate_helminth_tree_toresolve.nex")
```
```{r}
#write.csv(tre_dat, file = "ungulate_helminth_tree_table.csv", row.names = F)
```

Finally, I manually **resolved the tree**. I did this in two steps. First, I rearranged the major clades within each helminth group (i.e. an order level phylogeny). Then, I specifically looked at the ungulate parasites to see if those groups could be resolved, which affects the mapping. 

To resolve the tree, I manually modified the nexus file and then ran the following code block to visually check the modifications.

```{r}
#tre_dat <- read.csv(file = "ungulate_helminth_tree_table.csv")
#tree <- read.nexus(file = "ungulate_helminth_tree_toresolve.nex")
#tree$edge.length <- NULL # remove branch lengths
#
#tree_plot <- tree_checking(tree, tre_dat)
#ggsave(filename = 'for_resolving.pdf', width = 500, height = 400, units = "mm")
```
