---
title: "Improve tree"
author: "Dan Benesh"
date: "February 27, 2019"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ape)
library(castor) # for ancestral state mapping
library(ggtree)
library(RColorBrewer)
library(phytools)
library(geiger)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE)
theme_set(new = theme_bw())
```

```{r get tree}
tree <- read.nexus(file = "ungulate_helminth_tree.nex")
tree$edge.length <- NULL # remove branch lengths
```

```{r get data}
tre_dat <- read.csv(file = "ungulate_helminth_tree_table.csv")
```

```{r}
source("tree_checking_function.R")
```


Make the names in the data and tree consistent

```{r}
# in tree
fix_tip_names <- substr(tree$tip.label, start = 1, stop = regexpr(tree$tip.label, pattern = "_") - 1 )
tree$tip.label[which(fix_tip_names!="")] <- fix_tip_names[which(fix_tip_names!="")]

fix_node_names <- substr(tree$node.label, start = 1, stop = regexpr(tree$node.label, pattern = "_") - 1 )
tree$node.label[which(fix_node_names!="")] <- fix_node_names[which(fix_node_names!="")]

# in data
fix_data_names <- substr(tre_dat$otl_name, start = 1, stop = regexpr(tre_dat$otl_name, pattern = "_") - 1 )
tre_dat$otl_name_red <- tre_dat$otl_name
tre_dat$otl_name_red[which(fix_data_names!="")] <- fix_data_names[which(fix_data_names!="")]

rm(fix_data_names, fix_node_names, fix_tip_names)
```


```{r}
tree_plot <- tree_checking(tree, tre_dat)

ggsave(filename = 'for_resolving.pdf', width = 500, height = 400, units = "mm")
```

## Add missing tips

OK, now we want to add the following nematode tips to the tree: Ancylostomatidae, Ascarididae, Trichostrongylidae.

```{r}
tree <- compute.brlen(tree, power = 1 ) # add branch lengths, so taxa can be add along branch, not just at node

# add taxa to branch below internal node
tree <- bind.tip(tree, tip.label = "Ancylostomatidae", 
                 position = 0.0001,
                 where = length(tree$tip.label) + which(tree$node.label == "mrcaott46769ott67070"))
tre_dat$in_tree[which(tre_dat$family == 'Ancylostomatidae')] <- TRUE

# add to tip
tree <- bind.tip(tree, tip.label = "Ascarididae", 
                 where = which(tree$tip.label == "Toxocaridae"))
tre_dat$in_tree[which(tre_dat$family == 'Ascarididae')] <- TRUE


tree <- bind.tip(tree, tip.label = "Trichostrongylidae", 
                 where = which(tree$tip.label == "Haemonchidae"))
tre_dat$in_tree[which(tre_dat$family == 'Trichostrongylidae')] <- TRUE
```


Then we want to add the following trematode tips to the tree: Balanorchiidae, Choerocotyloididae, Zygocotylidae.

```{r}
filter(tre_dat, grepl(otl_name, pattern = "Choer"))
```


```{r}
# add taxa to internal node
tree <- bind.tip(tree, tip.label = "Balanorchiidae", 
                 where = length(tree$tip.label) + which(tree$node.label == "Paramphistomoidea"))
tre_dat$in_tree[which(tre_dat$family == 'Balanorchiidae')] <- TRUE

tree <- bind.tip(tree, tip.label = "Choerocotylidae", 
                 where = length(tree$tip.label) + which(tree$node.label == "Paramphistomoidea"))
tre_dat$in_tree[which(tre_dat$otl_name == 'Choerocotylidae')] <- TRUE

tree <- bind.tip(tree, tip.label = "Zygocotylidae", 
                 where = length(tree$tip.label) + which(tree$node.label == "Paramphistomoidea"))
tre_dat$in_tree[which(tre_dat$family == 'Zygocotylidae')] <- TRUE
```


## Double the taxa in cases where multiple strategies used

Which taxa? 
Syngamidae - int host, hatch and eaten
Ascarididae - transplacental, egg ingestion
Strongyloididae - transplacental, penetration

```{r}
filter(tre_dat, grepl(family, pattern = 'Strongylo'))
```

```{r}
# add to tip
tree <- bind.tip(tree, tip.label = "Syngamidae2", 
                 where = which(tree$tip.label == "Syngamidae"))

tree <- bind.tip(tree, tip.label = "Strongyloididae2", 
                 where = which(tree$tip.label == "Strongyloididae"))

tree <- bind.tip(tree, tip.label = "Ascarididae2", 
                 position = 0.0001,
                 where = getMRCA(tree, tip = c("Ascarididae", "Toxocaridae")))

tree <- bind.tip(tree, tip.label = "Toxocaridae2", 
                 where = which(tree$tip.label == "Ascarididae2"))

```

```{r}
# update data
syg2 <- data.frame(family = 'Syngamidae2', group = 'Nematoda', ung_inf = TRUE,
                   otl_name = 'Syngamidae2', trans_route = 'larvae on herbage', 
                   in_tree = TRUE, otl_name_red = 'Syngamidae2')

str2 <- data.frame(family = 'Strongyloididae2', group = 'Nematoda', ung_inf = TRUE,
                   otl_name = 'Strongyloididae2', trans_route = 'no free propagule', 
                   in_tree = TRUE, otl_name_red = 'Strongyloididae2')

asc2 <- data.frame(family = 'Ascarididae2', group = 'Nematoda', ung_inf = TRUE,
                   otl_name = 'Ascarididae2', trans_route = 'no free propagule', 
                   in_tree = TRUE, otl_name_red = 'Ascarididae2')

tox2 <- data.frame(family = 'Toxocaridae2', group = 'Nematoda', ung_inf = TRUE,
                   otl_name = 'Toxocaridae2', trans_route = 'no free propagule', 
                   in_tree = TRUE, otl_name_red = 'Toxocaridae2')

tre_dat <- bind_rows(tre_dat, syg2, str2, asc2, tox2)
rm(asc2, str2, syg2, tox2)
```


## Tidy up tree

Next, I'll remove the clades of free-living flatworms

```{r}
rem_platy <- tips(tree, length(tree$tip.label) + which(tree$node.label == 'Rhabditophora'))
rem_platy2 <- tips(tree, length(tree$tip.label) + which(tree$node.label == 'Catenulida'))
```

```{r}
# drop tips
tree <- drop.tip(tree, tip = rem_platy)
tree <- drop.tip(tree, tip = rem_platy2)
# revise data
tre_dat$in_tree[which(tre_dat$otl_name_red %in%  c(rem_platy, rem_platy2))] <- FALSE
```

```{r}
tree$edge.length <- NULL # remove branch lengths
tree_plot <- tree_checking(tree, tre_dat)
tree_plot
ggsave(filename = 'for_resolving.pdf', width = 500, height = 400, units = "mm")
```

```{r}
write.nexus(tree, file = "ungulate_helminth_tree_toresolve.nex")
```

```{r}
write.csv(tre_dat, file = "ungulate_helminth_tree_table.csv", row.names = F)
```



# Mess around with resolving tree

```{r}
tree <- read.nexus(file = "ungulate_helminth_tree_toresolve.nex")
tree$edge.length <- NULL # remove branch lengths
```

```{r}
tree_plot <- tree_checking(tree, tre_dat)

ggsave(filename = 'for_resolving.pdf', width = 500, height = 400, units = "mm")
```

