---
title: "Map states onto trees"
author: "Dan Benesh"
output: html_document
---

Here I want to determine how often different transmission strategies to ungulate hosts have evolved. To that end, I want to map anceestral states to see 1) how often ungulates have been invaded and 2) what routes were taken to infect them.

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ape)
library(castor) # for ancestral state mapping
library(ggtree)
library(RColorBrewer)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE)
theme_set(new = theme_bw())
```

```{r get tree}
tree <- read.nexus(file = "ungulate_helminth_tree_resolved.nex")
tree$edge.length <- NULL # remove branch lengths
```

```{r get data}
tre_dat <- read.csv(file = "ungulate_helminth_tree_table.csv")
```

Here is the tree of helminth families, acquired from the open tree of life. Some parts of it need to be resolved using phylogenies that are not included in the open tree.

```{r}
plot(tree, cex = 0.5, show.tip.label = F, show.node.label = T)
```

```{r}
# bifurcating produces slightly different mapping - better if non-random, i.e. I manually resolve tree
# tree <- compute.brlen(tree)
# tree <- multi2di(tree)
# tree$edge.length[which(tree$edge.length == 0)] <- 0.0001
```

I reduced the dataset of family names to just include the ones in the tree.

```{r}
# reduce to just that in tree
tre_dat <- filter(tre_dat, in_tree)

# which OTL names not in tree
uniq_otl_names <- unique(tre_dat$otl_name_red)
#tree$tip.label[!(tree$tip.label %in% uniq_otl_names)] # none after fixing group names in get_tree.rmd
```

There are more rows in data than tip labels. This is because some families were synonomized, resulting in several rows in the data with the same family (otl) name.

```{r}
group_by(tre_dat, otl_name)%>%
  summarize(n = n())%>%
  filter(n>1)
```

For those with multiple OTL entries for a given OTL name, I just took the first name.

```{r}
rm(tre_dat_red)
for(otl in uniq_otl_names){
  # if otl just
  dx <- filter(tre_dat, otl_name_red == otl)
  if(dim(dx)[1] > 1) {
    # if more than one entry for otl name, just take first
    print
    dx <- dx[1,]
  }
  
  if(exists('tre_dat_red')) {
    tre_dat_red <- rbind(tre_dat_red, dx)
  } else {
    tre_dat_red <- dx
  }
}
rm(otl, dx)
```

```{r}
# put data in same order as tree tip labels
mv <- match(tree$tip.label, tre_dat_red$otl_name_red)
tre_dat_red <- tre_dat_red[mv,]
rm(mv)
```

After making the trait data and tree consistent regarding names and order, I can start mapping. First things first, make a table showing the distribution of tip states.
```{r}
sort(table(tre_dat_red$trans_route), decreasing = T)
```


I am going to keep things simple and use a parsimony algorithm. It is fast and requires few assumptions about transition costs.

```{r}
# PARSIMONY
states_int <- map_to_state_space(tre_dat_red$ung_inf) # for the castor functions, trait has to be integer (faster)
ac1 <- asr_max_parsimony(tree, states_int$mapped_states, Nstates = 2)
```

```{r}
# palette for two states
mypalette <- brewer.pal(n = states_int$Nstates, "Set1")[1:2]
```

```{r}
tree_df <- fortify(tree) # make a df
```

```{r}
# ancestral state reconstructions to add to tree
# if states given same probability take 1st one (not ungulate parasite)
node_states <- apply(ac1$ancestral_likelihoods, 1, function(x) which(x == max(x))[1])
node_states <- unlist(node_states)

# combine tip and branch states, add them to tree df (correct order)
state <- c(states_int$mapped_states, node_states)
tree_df$state <- state
tree_df <- mutate(tree_df, 
                  state = if_else(state == 1, "not ungulate parasite", "ungulate parasite"))
```

After running the parsimony algorithm, let's plot the output, with ancestral states represented by colors on the tree.

```{r}
# plot tree
p <- ggtree(tree_df) + 
  geom_tree(aes(color = state)) + 
  geom_tiplab(data = filter(tree_df, state == 'ungulate parasite'),
              size = 1.5, color = mypalette[2]) +
  xlim(0, max(tree_df$x) + 0.09) +
  theme(legend.position = c(0.5, 1), 
        legend.justification = c(1,1),
        legend.title = element_blank(),
        legend.text = element_text(size = 14))
p
```
Evolutionary invasions of ungulates has occurred multiple times throughout the tree. This is no surprise; ungulates harbor a diverse and phylogenetically distant group of parasitic worms.
```{r}
# # plot tree with all tip labels
# p <- ggtree(tree_df) + 
#   geom_tree(aes(color = state)) + 
#   geom_tiplab(aes(color = state), size = 2) +
#   xlim(0, max(tree_df$x) + 0.09) +
#   scale_color_manual(values = mypalette) +
#   theme(legend.position = c(0.5, 1), 
#         legend.justification = c(1,1),
#         legend.title = element_blank(),
#         legend.text = element_text(size = 14))
```

```{r}
# plot with node labels
# p + geom_text2(aes(subset=!isTip, label=node), hjust=-.3)
```

Let's highlight the spots on the tree where transitions occured.

```{r}
# reduce to just tree structure and states
trans_states <- select(tree_df, child_node = node, parent_node = parent, child_state = state)
# self join to get states for both parent and child side by side
trans_states <- left_join(trans_states, 
                           select(trans_states, child_node, parent_state = child_state),
                           by = c("parent_node" = "child_node") ) 
# only cases where parent and child have different states
trans_states <- filter(trans_states, child_state != parent_state)%>%
  filter(child_state == "ungulate parasite")
#filter(tree_df, node %in% trans_statesx$child_node) # check if transitions correctly identified - looks good
```

How many transitions occurred?
```{r}
length(unique(trans_states$parent_node))
```

```{r}
# add transitions to plotting df
tree_df <- mutate(tree_df, trans_node = if_else(node %in% trans_states$parent_node, TRUE, FALSE))

# re-plot with points
p <- ggtree(tree_df, aes(color = state), size = 0.7, layout = 'circular') + 
  geom_tiplab2(hjust = -0.1, size = 1.25) +
  geom_nodepoint(data = filter(tree_df, trans_node), 
                 shape = 19, size = 2, color = mypalette[2], alpha = 0.75) +
  scale_color_manual(values = mypalette) +
  theme(legend.position = c(0.2, 0.95), 
        legend.justification = c(1,1),
        legend.title = element_blank(),
        legend.text = element_text(size = 15))
p
```

We can write this tree to file.

```{r}
ggsave(filename = 'transitions_to_ung_parasitism.pdf', width = 297, height = 210, units = "mm")
```

Now let's split the 'ungulate parasite' group into multiple groups, depending on how they infect ungulates. First, we define three trophic transmission strategies:
1. eggs ingested
2. free larvae ingested
3. intermediate host ingested

And then 3 additional non-trophic strategies:
4. active larvae, seek/penetrate
5. vector transmission
6. no free propagule - transmammary, transplacental or cannabilistic transmission

Then we re-map these onto the tree with parsimony.

```{r}
tre_dat_red <- mutate(tre_dat_red, trans_route = factor(trans_route, 
                                                        levels = c("not ungulate parasites",
                                                                   "eggs ingested",
                                                                   "larvae on herbage",
                                                                   "intermediate host eaten",
                                                                   "larvae seek/penetrate",
                                                                   "vector-transmission",
                                                                   "no free propagule")))
```

```{r}
# PARSIMONY
states_int <- as.integer(tre_dat_red$trans_route)
ac2 <- asr_max_parsimony(tree, states_int, Nstates = max(states_int))
```

```{r}
# palette for seven states
mypalette <- brewer.pal(n = max(states_int), "Set1")
```

```{r}
tree_df <- fortify(tree) # make a df
```

```{r}
# ancestral state reconstructions to add to tree
# if states given same probability take 1st one
node_states <- apply(ac2$ancestral_likelihoods, 1, function(x) which(x == max(x))[1])
node_states <- unlist(node_states)

# combine tip and branch states, add them to tree df (correct order)
state <- c(states_int, node_states)
tree_df$state <- levels(tre_dat_red$trans_route)[state]
```

```{r}
# plot tree
# p <- ggtree(tree_df, aes(color = state), size = 0.7, layout = 'circular') + 
#   geom_tiplab2(hjust = -0.1, size = 1.25) +
#   scale_color_manual(values = c(mypalette[1:5], 'light gray', mypalette[6])) +
#   theme(legend.position = c(0.1, 1), 
#         legend.justification = c(1,1),
#         legend.title = element_blank(),
#         legend.text = element_text(size = 15))
# p
```

```{r}
# reduce to just tree structure and states
trans_states2 <- select(tree_df, child_node = node, parent_node = parent, child_state = state)
# self join to get states for both parent and child side by side
trans_states2 <- left_join(trans_states2, 
                           select(trans_states2, child_node, parent_state = child_state),
                           by = c("parent_node" = "child_node") ) 

# add states to tree df - important for plotting
tree_df <- left_join(tree_df, 
                     select(trans_states2, child_node, parent_node, child_state, parent_state),
                     by = c("node" = "child_node"))
tree_df <- select(tree_df, -state) # use more useful columns - child and parent states

# id transition nodes
trans_states2 <- filter(trans_states2, child_state != parent_state)%>%
  filter(child_state != "not ungulate parasites")

tree_df <- mutate(tree_df, trans_node = if_else(node %in% trans_states2$parent_node, TRUE, FALSE))

# only cases where parent and child have different states
#filter(tree_df, node %in% trans_states2$child_node) # check if transitions correctly identified - looks good
```

```{r}
filter(trans_states2, !(parent_node %in% trans_states$parent_node))
```

```{r}
# re-plot with points
p <- ggtree(tree_df, aes(color = child_state), size = 0.7, layout = 'circular') +
  geom_tiplab2(hjust = -0.1, size = 1.25) +
  scale_color_manual(values = c(mypalette[1:5], 'light gray', mypalette[6])) +
  geom_nodepoint(data = filter(tree_df, trans_node), aes(color = parent_state),
                 shape = 19, size = 2, 
                 #color = "gray", 
                 alpha = 0.75) +
  theme(legend.position = c(0.1, 1),
        legend.justification = c(1,1),
        legend.title = element_blank(),
        legend.text = element_text(size = 15))

p
```

The plot shows the transitions.

```{r}
ggsave(filename = 'transitions_to_ung_trans_routes.pdf', width = 297, height = 210, units = "mm")
```
