---
title: "Getting taxonomic tree to map invasion of ungulates"
author: "Dan Benesh"
output: html_document
---

In this script, I get a tree from the OTL that will be used to map ungulate parasitism across helminths. I want to know how often these hosts have been invaded and which are most commonly strategies used to do so.

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(taxize)
library(rotl)
library(ape)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE)
theme_set(new = theme_bw())
```

First, I make a list of helminth families in which ungulate parasitism occurs. These were compiled by Jimmy Chubb.

```{r}
# list of helminth fams in ungulates - from MS
mon_ung_fams <- "Polystomatidae"
ces_ung_fams <- c('Diphyllobothridae', 'Anoplocephalinae', 'Linstowiinae', 'Thysanosomatinae', 'Inermicapsiferinae',
              'Davaineidae', 'Taeniidae')
tre_ung_fams <- c('Hasstilesiidae', 'Schistosomatidae', 'Echinostomatidae', 'Fasciolidae', 'Paramphistomidae',
              'Balanorchiidae', 'Brumptiidae', 'Choerocotyloididae', 'Cladorchiidae', 'Gastrodiscidae',
              'Gastrothylacidae', 'Olveriidae', 'Stephanopharyngidae', 'Zygocotylidae', 'Notocotylidae',
              'Opisthorchiidae', 'Heterophyidae', 'Dicrocoeliidae', 'Paragonimidae')
nem_ung_fams <- c('Capillariidae', 'Robertdollfusidae', 'Trichinellidae', 'Trichuridae', 'Ancylostomatidae',
               'Ascarididae', 'Chabertiidae', 'Cooperiidae', 'Dictyocaulidae', 'Filariidae', 'Gnathostomatidae',
               'Gongylonematidae', 'Habronematidae', 'Haemonchidae', 'Metastrongylidae', 'Molineidae',
               'Onchocercidae', 'Oxyuridae', 'Protostrongylidae', 'Spirocercidae', 'Strongylidae', 'Strongyloididae',
               'Syngamidae', 'Thelaziidae', 'Trichostrongylidae')
aca_ung_fams <- c('Oligacanthorhynchidae','Echinorhynchidae')
```

To be able to map ancestral states, we also need to know in which helminths ungulate parasitism does not occur. In other words, we need to get the related taxa that are not parasites of ungulates. For this, I queried COL and ITIS with `taxize` to find all families in Acanthocephala, Nematoda, and Platyhelmintha.

Starting with the acanths, who have only sporadically been found in ungulates. Both families with records from ungulates were in the taxonomy databases.

```{r}
# from COL
aca_col_id <- "29cf5a3a0cad337a0216edca32bab745" # fetched before
aca_col_fams <- downstream(aca_col_id, downto = 'family', db = 'col')
aca_col_fams <- aca_col_fams$`29cf5a3a0cad337a0216edca32bab745`$childtaxa_name
# from ITIS
aca_itis_id <- 64238 # looked it up online
aca_itis_fams <- downstream(aca_itis_id, downto = 'family', db = 'itis')
aca_itis_fams <- aca_itis_fams$`64238`$taxonname

#sum(aca_col_fams %in% aca_itis_fams)/length(aca_col_fams) # percent COL in ITIS
#sum(aca_ung_fams %in% aca_col_fams)/length(aca_ung_fams) # percent helminth fams in COL
#sum(aca_ung_fams %in% aca_itis_fams)/length(aca_ung_fams) # percent helminth fams in ITIS
```

Twenty-five nematode families have been recorded in ungulate hosts. About 80% of these were in COL, but only about 50% in ITIS.

```{r}
# get list of all nematode families
# from COL
nem_col_id <- get_colid(sciname = "Nematoda")
nem_col_fams <- downstream(nem_col_id, downto = 'family', db = 'col')
nem_col_fams <- nem_col_fams$`562405d59c95f93423494a00939ae2ff`$childtaxa_name
# from ITIS
nem_itis_id <- get_tsn(searchterm = "Nematoda")
nem_itis_fams <- downstream(nem_itis_id, downto = 'family', db = 'itis')
nem_itis_fams <- nem_itis_fams$`59490`$taxonname

#sum(nem_col_fams %in% nem_itis_fams)/length(nem_col_fams) # percent COL in ITIS
#sum(nem_ung_fams %in% nem_col_fams)/length(nem_ung_fams) # percent helminth fams in COL
#sum(nem_ung_fams %in% nem_itis_fams)/length(nem_ung_fams) # percent helminth fams in ITIS
```

Each of the three groups of parasitic playhelminths (Monogenea, Trematoda, Cestoda) have members infecting ungulates. But only about 40 to 60% of these 27 families have a match in the two taxonomy databases. The discrepancy suggests either Jimmy or the databases are using outdated family names, which is a common issue.

```{r}
# get list of all platy families
# from COL
pla_col_id <- get_colid(sciname = "Platyhelminthes")
pla_col_fams <- downstream(pla_col_id, downto = 'family', db = 'col')
pla_col_fams <- pla_col_fams$`7253b0d111fd5eb0c6a9856e0d8d1cad`$childtaxa_name
# from ITIS
pla_itis_id <- get_tsn(searchterm = "Platyhelminthes")
pla_itis_fams <- downstream(pla_itis_id, downto = 'family', db = 'itis')
pla_itis_fams <- pla_itis_fams$`53963`$taxonname

pla_ung_fams <- c(mon_ung_fams, tre_ung_fams, ces_ung_fams)
sum(pla_col_fams %in% pla_itis_fams)/length(pla_col_fams) # percent COL in ITIS
sum(pla_ung_fams %in% pla_col_fams)/length(pla_ung_fams) # percent helminth fams in COL
sum(pla_ung_fams %in% pla_itis_fams)/length(pla_ung_fams) # percent helminth fams in ITIS
```

Given that the helminth ungulate families and the taxonomy datasets are not completely congruent, I chose to Combine all three lists of family names to query the OTL. Some might be redundant or invalid, but this should give the largest, most complete tree with which to map ancestral states.

```{r}
# combine theme all
aca_fams_all <- unique(c(aca_ung_fams, aca_col_fams, aca_itis_fams))
nem_fams_all <- unique(c(nem_ung_fams, nem_col_fams, nem_itis_fams))
pla_fams_all <- unique(c(pla_ung_fams, pla_col_fams, pla_itis_fams))
```

```{r}
# search OTL db
aca_search <- tnrs_match_names(names = aca_fams_all, context_name = "Animals")
nem_search <- tnrs_match_names(names = nem_fams_all, context_name = "Nematodes")
pla_search1 <- tnrs_match_names(names = pla_fams_all[1:240], context_name = "Platyhelminthes")
pla_search2 <- tnrs_match_names(names = pla_fams_all[241:length(pla_fams_all)], context_name = "Platyhelminthes")
```

```{r}
# which ones are in tree
aca_ott_in_tree <- ott_id(aca_search)[is_in_tree(ott_id(aca_search))]
nem_ott_in_tree <- ott_id(nem_search)[is_in_tree(ott_id(nem_search))]
pla_ott <- c(ott_id(pla_search1), ott_id(pla_search2))
pla_ott_in_tree <- pla_ott[is_in_tree(pla_ott)]
# attributes erased by appending lists together - re-add so subtree function works
attr(pla_ott_in_tree, 'class') <- c("otl_ott_id", "list")
```

```{r}
# get subtrees
aca_tree <- tol_induced_subtree(ott_ids = aca_ott_in_tree, label_format = 'name')
nem_tree <- tol_induced_subtree(ott_ids = nem_ott_in_tree, label_format = 'name')
pla_tree <- tol_induced_subtree(ott_ids = pla_ott_in_tree, label_format = 'name')
```
```{r}
taxa_searched <- length(aca_fams_all) + length(nem_fams_all) + length(pla_fams_all)
taxa_in_tree <- length(aca_tree$tip.label) + length(nem_tree$tip.label) + length(pla_tree$tip.label)
```

Of `r length(taxa_searched)` families queried, `r length(taxa_in_tree)` were included in the tree. All acanth families of ungulates were in the tree. Here are the nematodes families of ungulates that were not included.

```{r}
# ungulate nem families not in nem tree
nem_ung_fams[which(!(nem_ung_fams %in% nem_tree$tip.label))]
```

And here are the platyhelminth families infecting ungulates not in the tree.

```{r}
pla_ung_fams[which(!(pla_ung_fams %in% pla_tree$tip.label))]
```

Some of these do not need to be added to the tree, such as the subfamilies of Anoplocephalidae (Linstowiinae, Thysanosomatinae, Inermicapsiferinae), because a similar type of transmission route applies to all these families (mite transmission). For other cases, a closer look at the tree is warranted.
Bind trees together to make full tree

```{r}
aca_tree$root.edge <- .2 # root edge needed to bind trees not exactly at root
nem_tree$root.edge <- .2
pla_tree$root.edge <- .2
comb_tree <- bind.tree(pla_tree, aca_tree, position = .2)
comb_tree <- bind.tree(comb_tree, nem_tree, position = .2)
```


```{r}
# make dataset (still need to add route)
helminth_ungs <- data.frame(family = c(aca_fams_all, nem_fams_all, pla_fams_all))
helminth_ungs <- mutate(helminth_ungs, group = if_else(family %in% aca_fams_all, "Acanthocephala",
                                                       if_else(family %in% pla_fams_all, "Platyhelminthes", 
                                                               "Nematoda")),
                        ung_inf = family %in% c(aca_ung_fams, nem_ung_fams, pla_ung_fams),
                        in_tree = family %in% comb_tree$tip.label)
```

Taxa to add to tree, manually?

```{r}
filter(helminth_ungs, ung_inf, !in_tree)
```

```{r}
# write files
write.csv(helminth_ungs, file = "ungulate_helminth_tree_table.csv", row.names = F)
write.nexus(comb_tree, file = "ungulate_helminth_tree.nex")
```




```{r}
# list of helminth taxa in ungulates - from London NHM DB, need ungulates for cross-checking
```
